% The contents of this file is 
% Copyright (c) 2009-  Charles R. Severance, All Righs Reserved

\chapter{Regular expressions}

So far we have been reading through files, looking for patterns and extracting various
bits of lines that we find interesting.  We have been using string methods like {\tt split}
and {\tt find} and using lists and string slicing to extract portions of the lines.
\index{regular expressions}
\index{regex}
\index{re module}

This task of searching and extracting is so common that Python has a very powerful library
called {\bf regular expressions} that handles many of these tasks quite elegantly.  The
reason we have not introduced regular expressions earlier in the book is because while they
are very powerful, they are a little complicated and their syntax takes some getting used to. 

Regular expressions are almost their own little programming language for searching and parsing
strings.  As a matter of fact, entire books have been written on the topic of regular expressions.
In this chapter, we will only cover the basics of regular expressions.  For more detail on regular
expressions, see:

\url{http://en.wikipedia.org/wiki/Regular_expression}

\url{https://docs.python.org/2/library/re.html}

The regular expression library {\tt re} must be imported into your program before you can use it.
The simplest use of the regular expression library is the {\tt search()} function.  The following
program demonstrates a trivial use of the search function.
\index{regex!search}

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    if re.search('From:', line) :
        print line
\end{verbatim}
\afterverb
%
We open the file, loop through each line, and use the regular expression {\tt search()} to
only print out lines that contain the string ``From:''.   This program does not use the real
power of regular expressions, since we could have just as easily used {\tt line.find()} to
accomplish the same result.
\index{string!find}

The power of the regular expressions comes when we add special characters to the search string
that allow us to more precisely control which lines match the string.  Adding these special
characters to our regular expression allow us to do sophisticated matching and extraction while
writing very little code.

For example, the caret character is used in regular 
expressions to match ``the beginning'' of a line.
We could change our program to only match 
lines where ``From:'' was at the beginning of the line as follows:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    if re.search('^From:', line) :
        print line
\end{verbatim}
\afterverb
%
Now we will only match lines that {\em start with} the string ``From:''.  This is still a very
simple example that we could have done equivalently with the {\tt startswith()} method from
the string library.  But it serves to introduce the notion that regular expressions contain
special action characters that give us more control as to what will match the regular expression.
\index{string!startswith}

\section{Character matching in regular expressions}

There are a number of other special characters that let us build even more powerful regular
expressions.  The most commonly used special character is the period or full stop, which matches
any character.
\index{wild card}
\index{regex!wild card}

In the following example, the regular expression ``F..m:'' would match any of the strings
``From:'', ``Fxxm:'', ``F12m:'', or ``F!@m:'' since the period characters in the regular
expression match any character.

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    if re.search('^F..m:', line) :
        print line
\end{verbatim}
\afterverb
%
This is particularly powerful when combined with the ability to indicate that a character can
be repeated any number of times using the ``*'' or ``+'' characters in your regular expression.
These special characters mean that instead of matching a single character in the search string,
they match zero-or-more characters (in the case of the asterisk) or one-or-more of the characters
(in the case of the plus sign).

We can further narrow down the lines that we match using a repeated {\bf wild card} character in
the following example:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    if re.search('^From:.+@', line) :
        print line
\end{verbatim}
\afterverb
%
The search string ``\verb"^"From:.+@'' will successfully match lines that start with ``From:'',
followed by one or more characters (``.+''), followed by an at-sign.  So this will match the
following line:

\beforeverb
\begin{alltt}
{\bf From:}\underline{ stephen.marquard}{\bf @}uct.ac.za
\end{alltt}
\afterverb

You can think of the ``.+'' wildcard as expanding to match all the characters between the 
colon character and the at-sign.  

\beforeverb
\begin{alltt}
{\bf From:}\underline{.+}{\bf @}
\end{alltt}
\afterverb

It is good to think of the plus and asterisk characters as ``pushy''.  For example, the following
string would match the last at-sign in the string as the ``.+'' pushes outwards, as shown below:

\beforeverb
\begin{alltt}
{\bf From:}\underline{ stephen.marquard@uct.ac.za, csev@umich.edu, and cwen}{\bf @}iupui.edu
\end{alltt}
\afterverb

It is possible to tell an asterisk or plus sign not to be so ``greedy'' by adding 
another character.   See the detailed documentation for information on turning off the 
greedy behavior.
\index{greedy}

\section{Extracting data using regular expressions}

If we want to extract data from a string in Python we can use the {\tt findall()} method to extract
all of the substrings which match a regular expression.  Let's use the example of wanting to extract
anything that looks like an email address from any line regardless of format.  For example, we want
to pull the email addresses from each of the following lines:

\beforeverb
\begin{verbatim}
From stephen.marquard@uct.ac.za Sat Jan  5 09:14:16 2008
Return-Path: <postmaster@collab.sakaiproject.org>
          for <source@collab.sakaiproject.org>;
Received: (from apache@localhost)
Author: stephen.marquard@uct.ac.za
\end{verbatim}
\afterverb
%
We don't want to write code for each of the types of lines, splitting and slicing differently for each
line.  This following program uses {\tt findall()} to find the lines with email addresses in them and
extract one or more addresses from each of those lines.
\index{findall}
\index{regex!findall}

\beforeverb
\begin{verbatim}
import re
s = 'Hello from csev@umich.edu to cwen@iupui.edu about the meeting @2PM'
lst = re.findall('\S+@\S+', s)
print lst
\end{verbatim}
\afterverb
%
The {\tt findall()} method searches the string in the second argument and returns a list of
all of the strings that look like email addresses.   We are using a two-character sequence 
that matches a non-whitespace character ({\textbackslash}S). 

The output of the program would be:

\beforeverb
\begin{verbatim}
['csev@umich.edu', 'cwen@iupui.edu']
\end{verbatim}
\afterverb
%
Translating the regular expression, we are looking for substrings that have at least one
non-whitespace character, followed by an at-sign, followed by at least one more non-whitespace
character.  The ``{\textbackslash}S+'' matches as many non-whitespace characters as possible.

The regular expression would match twice (csev@umich.edu and cwen@iupui.edu), but it would not
match the string ``@2PM'' because there are no non-blank characters {\em before} the at-sign.  
We can use this regular expression in a program to read all the lines in a file and print out
anything that looks like an email address as follows:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    x = re.findall('\S+@\S+', line)
    if len(x) > 0 :
        print x
\end{verbatim}
\afterverb
%
We read each line and then extract all the substrings that match our regular expression.
Since {\tt findall()} returns a list, we simply check if the number of elements in our returned
list is more than zero to print only lines where we found at least one substring that looks
like an email address.

If we run the program on {\tt mbox.txt} we get the following output:

\beforeverb
\begin{verbatim}
['wagnermr@iupui.edu']
['cwen@iupui.edu']
['<postmaster@collab.sakaiproject.org>']
['<200801032122.m03LMFo4005148@nakamura.uits.iupui.edu>']
['<source@collab.sakaiproject.org>;']
['<source@collab.sakaiproject.org>;']
['<source@collab.sakaiproject.org>;']
['apache@localhost)']
['source@collab.sakaiproject.org;']
\end{verbatim}
\afterverb
%
Some of our email addresses have incorrect characters like ``\verb"<"'' or ``;'' at the beginning
or end.   Let's declare that we are only interested in the portion of the string that starts and
ends with a letter or a number.

To do this, we use another feature of regular expressions.  Square brackets are used to indicate a
set of multiple acceptable characters we are willing to consider matching.  In a sense, the
``{\textbackslash}S'' is asking to match the set of ``non-whitespace characters''.  Now we will be
a little more explicit in terms of the characters we will match.

Here is our new regular expression:

\beforeverb
\begin{verbatim}
[a-zA-Z0-9]\S*@\S*[a-zA-Z]
\end{verbatim}
\afterverb
%
This is getting a little complicated and you can begin to see why regular expressions are their own
little language unto themselves.  Translating this regular expression, we are looking for substrings
that start with a {\em single} lowercase letter, uppercase letter, or number ``[a-zA-Z0-9]'', followed
by zero or more non-blank characters (``{\textbackslash}S*''), followed by an at-sign, followed by zero
or more non-blank characters (``{\textbackslash}S*''), followed by an uppercase or lowercase letter.
Note that we switched from ``+'' to ``*'' to indicate zero or more non-blank characters since ``[a-zA-Z0-9]''
is already one non-blank character.   Remember that the ``*'' or ``+'' applies to the single character
immediately to the left of the plus or asterisk.
\index{regex!character sets(brackets)}

If we use this expression in our program, our data is much cleaner:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    x = re.findall('[a-zA-Z0-9]\S*@\S*[a-zA-Z]', line)
    if len(x) > 0 :
        print x
\end{verbatim}
\afterverb
%

\beforeverb
\begin{verbatim}
...
['wagnermr@iupui.edu']
['cwen@iupui.edu']
['postmaster@collab.sakaiproject.org']
['200801032122.m03LMFo4005148@nakamura.uits.iupui.edu']
['source@collab.sakaiproject.org']
['source@collab.sakaiproject.org']
['source@collab.sakaiproject.org']
['apache@localhost']
\end{verbatim}
\afterverb
%
Notice that on the ``source@collab.sakaiproject.org'' lines, our regular expression
eliminated two letters at the end of the string (``\verb">";'').  This is because when we
append ``[a-zA-Z]'' to the end of our regular expression, we are demanding that whatever
string the regular expression parser finds must end with a letter.   So when it sees the
``\verb">"'' after ``sakaiproject.org\verb">";'' it simply stops at the last ``matching''
letter it found (i.e., the ``g'' was the last good match).

Also note that the output of the program is a Python list that has a string as the single
element in the list.

\section{Combining searching and extracting}

If we want to find numbers on lines that start with the string ``X-'' such as:

\beforeverb
\begin{verbatim}
X-DSPAM-Confidence: 0.8475
X-DSPAM-Probability: 0.0000  
\end{verbatim}
\afterverb
%
we don't just want any floating-point numbers from any lines.  We only want to extract
numbers from lines that have the above syntax.

We can construct the following regular expression to select the lines:

\beforeverb
\begin{verbatim}
^X-.*: [0-9.]+
\end{verbatim}
\afterverb
%
Translating this, we are saying, we want lines that start with ``X-'', followed by zero or
more characters (``.*''), followed by a colon (``:'') and then a space.  After the space we are
looking for one or more characters that are either a digit (0-9) or a period ``[0-9.]+''.
Note that inside the square brackets, the period matches an actual period (i.e., it is not a
wildcard between the square brackets).

This is a very tight expression that will pretty much match only the lines we are interested
in as follows:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    if re.search('^X\S*: [0-9.]+', line) :
        print line
\end{verbatim}
\afterverb
%
When we run the program, we see the data nicely filtered to show 
only the lines we are looking for.

\beforeverb
\begin{verbatim}
X-DSPAM-Confidence: 0.8475
X-DSPAM-Probability: 0.0000
X-DSPAM-Confidence: 0.6178
X-DSPAM-Probability: 0.0000
\end{verbatim}
\afterverb
%
But now we have to solve the problem of extracting the numbers.  While it would be simple
enough to use {\tt split}, we can use another feature of regular expressions to both search
and parse the line at the same time.
\index{string!split}

Parentheses are another special character in regular expressions.  When you add parentheses
to a regular expression, they are ignored when matching the string. But when you are using
{\tt findall()}, parentheses indicate that while you want the whole expression to match,
you only are interested in extracting a portion of the substring that matches the regular
expression.  
\index{regex!parentheses}
\index{parentheses!regular expression}

So we make the following change to our program:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    x = re.findall('^X\S*: ([0-9.]+)', line)
    if len(x) > 0 :
        print x
\end{verbatim}
\afterverb
%
Instead of calling {\tt search()}, we add parentheses around the part of the regular expression
that represents the floating-point number to indicate we only want {\tt findall()} to give us
back the floating-point number portion of the matching string.

The output from this program is as follows:

\beforeverb
\begin{verbatim}
['0.8475']
['0.0000']
['0.6178']
['0.0000']
['0.6961']
['0.0000']
..
\end{verbatim}
\afterverb
%
The numbers are still in a list and need to be converted from strings to floating point, but we
have used the power of regular expressions to both search and extract the information we found
interesting.

As another example of this technique, if you look at the file there are a number of lines
of the form:

\beforeverb
\begin{verbatim}
Details: http://source.sakaiproject.org/viewsvn/?view=rev&rev=39772
\end{verbatim}
\afterverb
%
If we wanted to extract all of the revision numbers (the integer number at the end of these lines)
using the same technique as above, we could write the following program:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    x = re.findall('^Details:.*rev=([0-9]+)', line)
    if len(x) > 0:
        print x
\end{verbatim}
\afterverb
%
Translating our regular expression, we are looking for lines that start with ``Details:'',
followed by any number of characters (``.*''), followed by ``rev='', and then by one or
more digits.   We want to find lines that match the entire expression but we only want to
extract the integer number at the end of the line, so we surround ``[0-9]+'' with parentheses.  

When we run the program, we get the following output:

\beforeverb
\begin{verbatim}
['39772']
['39771']
['39770']
['39769']
...
\end{verbatim}
\afterverb
%
Remember that the ``[0-9]+'' is ``greedy'' and it tries to make as large a string of digits as
possible before extracting those digits.  This ``greedy'' behavior is why we get all five digits
for each number.  The regular expression library expands in both directions until it encounters a
non-digit, or the beginning or the end of a line.

Now we can use regular expressions to redo an exercise from earlier in the book where we were
interested in the time of day of each mail message.   We looked for lines of the form:

\beforeverb
\begin{verbatim}
From stephen.marquard@uct.ac.za Sat Jan  5 09:14:16 2008
\end{verbatim}
\afterverb
%
and wanted to extract the hour of the day for each line.  Previously we did this with two calls
to {\tt split}.  First the line was split into words and then we pulled out the fifth word and split
it again on the colon character to pull out the two characters we were interested in.

% Add a section on the notion of brittle code
While this worked, it actually results in pretty brittle code that is assuming the lines are nicely
formatted.  If you were to add enough error checking (or a big try/except block) to insure that your
program never failed when presented with incorrectly formatted lines, the code would balloon to 
10-15 lines of code that was pretty hard to read.

We can do this in a far simpler way with the following regular expression:

\beforeverb
\begin{verbatim}
^From .* [0-9][0-9]:
\end{verbatim}
\afterverb
%
The translation of this regular expression is that we are looking for lines that start with ``From ''
(note the space), followed by any number of characters (``.*''), followed by a space, followed by two
digits ``[0-9][0-9]'', followed by a colon character.  This is the definition of the kinds of lines
we are looking for.  

In order to pull out only the hour using {\tt findall()}, we add parentheses around the two digits
as follows:

\beforeverb
\begin{verbatim}
^From .* ([0-9][0-9]):
\end{verbatim}
\afterverb
%
This results in the following program:

\beforeverb
\begin{verbatim}
import re
hand = open('mbox-short.txt')
for line in hand:
    line = line.rstrip()
    x = re.findall('^From .* ([0-9][0-9]):', line)
    if len(x) > 0 : print x
\end{verbatim}
\afterverb
%
When the program runs, it produces the following output:

\beforeverb
\begin{verbatim}
['09']
['18']
['16']
['15']
...
\end{verbatim}
\afterverb
%
\section{Escape character}

Since we use special characters in regular expressions to match the beginning or end of 
a line or specify wild cards, we need a way to indicate that these characters are ``normal'' 
and we want to match the actual character such as a dollar sign or caret.

We can indicate that we want to simply match a character by prefixing that character 
with a backslash.  For example, we can find money amounts with the following regular
expression.

\beforeverb
\begin{verbatim}
import re
x = 'We just received $10.00 for cookies.'
y = re.findall('\$[0-9.]+',x)
\end{verbatim}
\afterverb
%
Since we prefix the dollar sign with a backslash, it actually matches the dollar sign
in the input string instead of matching the ``end of line'', and the rest of the regular
expression matches one or more digits or the period character.  {\em Note:} Inside 
square brackets, characters are not ``special''.   So when we say ``[0-9.]'', it really 
means digits or a period.    Outside of square brackets, a period is the ``wild-card'' 
character and matches any character.  Inside square brackets, the period is a period.

\section{Summary}

While this only scratched the surface of regular expressions, we have learned a bit about
the language of regular expressions.  They are search strings with special characters in them
that communicate your wishes to the regular expression system as to what defines ``matching''
and what is extracted from the matched strings.  Here are some of those special characters
and character sequences:

\verb"^" \newline
Matches the beginning of the line.

\$ \newline
Matches the end of the line.

. \newline
Matches any character (a wildcard).

{\textbackslash}s \newline
Matches a whitespace character.

{\textbackslash}S \newline
Matches a non-whitespace character (opposite of {\textbackslash}s).

* \newline
Applies to the immediately preceding character and indicates to match zero or more of the
preceding character(s).

*? \newline
Applies to the immediately preceding character and indicates to match zero or more of the
preceding character(s) in ``non-greedy mode''.

+ \newline
Applies to the immediately preceding character and indicates to match one or more of the 
preceding character(s).

+? \newline
Applies to the immediately preceding character and indicates to match one or more of the 
preceding character(s) in ``non-greedy mode''.

[aeiou] \newline
Matches a single character as long as that character is in the specified set.  In this example,
it would match ``a'', ``e'', ``i'', ``o'', or ``u'', but no other characters.

[a-z0-9] \newline
You can specify ranges of characters using the minus sign.  This example is a single character
that must be a lowercase letter or a digit.

[\verb"^"A-Za-z] \newline
When the first character in the set notation is a caret, it inverts the logic.  This example
matches a single character that is anything {\em other than} an uppercase or lowercase letter.

( ) \newline
When parentheses are added to a regular expression, they are ignored for the purpose of matching,
but allow you to extract a particular subset of the matched string rather than the whole string
when using {\tt findall()}.

{\textbackslash}b \newline
Matches the empty string, but only at the start or end of a word.

{\textbackslash}B \newline
Matches the empty string, but not at the start or end of a word.

{\textbackslash}d \newline
Matches any decimal digit; equivalent to the set [0-9].

{\textbackslash}D \newline
Matches any non-digit character; equivalent to the set [\verb"^"0-9].

\section{Bonus section for Unix users}

Support for searching files using regular expressions was built into the Unix operating system 
since the 1960s and it is available in nearly all programming languages in one form or another.

\index{grep}
As a matter of fact, there is a command-line program built into Unix 
called {\bf grep} (Generalized Regular Expression Parser) that does pretty much 
the same as the {\tt search()} examples in this chapter.  So if you have a 
Macintosh or Linux system, you can try the following commands in your command-line window.

\beforeverb
\begin{verbatim}
$ grep '^From:' mbox-short.txt
From: stephen.marquard@uct.ac.za
From: louis@media.berkeley.edu
From: zqian@umich.edu
From: rjlowe@iupui.edu
\end{verbatim}
\afterverb
%
This tells {\tt grep} to show you lines that start with the string ``From:'' in the file
{\tt mbox-short.txt}.   If you experiment with the {\tt grep} command a bit and read the
documentation for {\tt grep}, you will find some subtle differences between the regular
expression support in Python and the regular expression support in {\tt grep}.  As an example,
{\tt grep} does not support the non-blank character ``{\textbackslash}S'' so you will need to
use the slightly more complex set notation ``[\verb"^" ]'', which simply means match a 
character that is anything other than a space.

\section{Debugging}

Python has some simple and rudimentary built-in documentation that can be quite helpful if
you need a quick refresher to trigger your memory about the exact name of a particular method.
This documentation can be viewed in the Python interpreter in interactive mode.

You can bring up an interactive help system using {\tt help()}.

\beforeverb
\begin{verbatim}
>>> help()

Welcome to Python 2.6!  This is the online help utility.

If this is your first time using Python, you should definitely check out
the tutorial on the Internet at http://docs.python.org/tutorial/.

Enter the name of any module, keyword, or topic to get help on writing
Python programs and using Python modules.  To quit this help utility and
return to the interpreter, just type "quit".

To get a list of available modules, keywords, or topics, type "modules",
"keywords", or "topics".  Each module also comes with a one-line summary
of what it does; to list the modules whose summaries contain a given word
such as "spam", type "modules spam".

help> modules
\end{verbatim}
\afterverb
%
If you know what module you want to use, you can use the {\tt dir()} command to find the methods in the module as follows:

\beforeverb
\begin{verbatim}
>>> import re
>>> dir(re)
[.. 'compile', 'copy_reg', 'error', 'escape', 'findall', 
'finditer', 'match', 'purge', 'search', 'split', 'sre_compile', 
'sre_parse', 'sub', 'subn', 'sys', 'template']
\end{verbatim}
\afterverb
%
You can also get a small amount of documentation on a particular method using the dir command.

\beforeverb
\begin{verbatim}
>>> help (re.search)
Help on function search in module re:

search(pattern, string, flags=0)
    Scan through string looking for a match to the pattern, returning
    a match object, or None if no match was found.
>>> 
\end{verbatim}
\afterverb
%
The built-in documentation is not very extensive, but it can be helpful when you are in a hurry
or don't have access to a web browser or search engine.

\section{Glossary}

\begin{description}

\item[brittle code:]
Code that works when the input data is in a particular format but is prone to breakage
if there is some deviation from the correct format.  We call this ``brittle code'' 
because it is easily broken.

\item[greedy matching:]
The notion that the ``+'' and ``*'' characters in a regular expression expand outward to
match the largest possible string.
\index{greedy}
\index{greedy matching}

\item[grep:]
A command available in most Unix systems that searches through text files looking for lines
that match regular expressions. The command name stands for "Generalized Regular Expression Parser".
\index{grep}

\item[regular expression:]
A language for expressing more complex search strings.  A regular expression may contain
special characters that indicate that a search only matches at the beginning or end of a line
or many other similar capabilities.

\item[wild card:]
A special character that matches any character.  In regular expressions the wild-card
character is the period.
\index{wild card}

\end{description}

\section{Exercises}

\begin{ex}
Write a simple program to simulate the operation of the {\tt grep} command 
on Unix.  Ask the user to enter a regular expression and count the number
of lines that matched the regular expression:

\beforeverb
\begin{verbatim}
$ python grep.py
Enter a regular expression: ^Author
mbox.txt had 1798 lines that matched ^Author

$ python grep.py
Enter a regular expression: ^X-
mbox.txt had 14368 lines that matched ^X-

$ python grep.py
Enter a regular expression: java$
mbox.txt had 4218 lines that matched java$
\end{verbatim}
\afterverb
%
\end{ex}

\begin{ex}
Write a program to look for lines of the form

\verb"New Revision: 39772"

and extract the number from each of the lines using a regular expression
and the {\tt findall()} method.  Compute the average of the numbers and 
print out the average.

\beforeverb
\begin{verbatim}
Enter file:mbox.txt 
38549.7949721

Enter file:mbox-short.txt
39756.9259259
\end{verbatim}
\afterverb
%

\end{ex}

